Darwin's Dream: Unifying ecological and evolutionary change

### 2025-05-29 foundations_of_ecoevo.tex
Initialise response to reviewer comments, with current manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\textbf{Biological evolution is realised through the same mechanisms of
birth and death that underlie change in population density.} \textbf{The
deep interdependence between ecology and evolution is well-established,
but much theory in each discipline has been developed in isolation.}
\textbf{While recent work has accomplished eco-evolutionary synthesis, a
gap remains between the logical foundations of ecology and evolution.}
\textbf{We bridge this gap with a new equation that defines a summed
value for a characteristic across individuals in a population, from
which the fundamental equations of population ecology and evolutionary
biology (the Price equation) are derived.} \textbf{We thereby unify the
fundamental equations of population ecology and biological evolution
under a general framework.} \textbf{Our unification further demonstrates
the equivalence between mean population growth rate and evolutionary
fitness, shows how ecological and evolutionary change are reflected in
the first and second statistical moments of fitness, respectively, and
links this change to ecosystem function.} \textbf{Finally, we outline
how our proposed framework can be used to unify social evolution and
density-dependent population growth.}

@@

\hypertarget{non-technical-summary}{%
\section{Non-technical Summary}\label{non-technical-summary}}

Biologists have long known that the ecology of population size is
inextricably linked with the evolution of population traits.
Nevertheless, the foundational theory underlying ecological and
evolutionary change have been historically separated, with recent
attempts to bridge this theoretical divide using eco-evolutionary
modelling. We discover a new equation from first principles that
provides a universal and complete definition for eco-evolutionary
change. This equation thereby offers a unified framework for ecology and
evolution, which also recovers the foundational theory of ecology and
evolution as developed in isolation.

@@

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\protect\hyperlink{ref-Kitcher1993}{Kitcher
1993}), and its value is perhaps most evident in reconciling unconnected
models and revealing new and unexpected empirical predictions. In
theoretical physics, Newton's unification of celestial and terrestrial
motion, and Maxwell's unification of electricity and magnetism, are
among the most important advances in scientific history. Likewise, in
the life sciences, the modern synthesis of natural selection and
Mendelian inheritance revolutionised how mechanisms underlying evolution
and genetics are understood and applied
(\protect\hyperlink{ref-Gayon2019}{Gayon and Huneman 2019}). The analogy
to physics is imperfect because evolution encompasses non-biological
systems with non-genetic inheritance (e.g.,
\protect\hyperlink{ref-Price1995}{Price 1995};
\protect\hyperlink{ref-Shennan2011}{Shennan 2011};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Lewens2015}{Lewens 2015}), but to illustrate our
point, we restrict our focus to biological evolution. Following the
modern synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

@@

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (e.g., \protect\hyperlink{ref-Lion2018}{Lion
2018}; \protect\hyperlink{ref-Patel2018}{Patel, Cortez, and Schreiber
2018}; \protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
foundational assumptions and equations of ecology and evolution in these
models remain mostly separated conceptually. But like the mechanisms
driving evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\protect\hyperlink{ref-Connor2004}{Connor and Hartl 2004}). Most
notably, change in population density and phenotype are both caused by
individual birth and death (\protect\hyperlink{ref-Turchin2001}{Turchin
2001}). This should allow biologists to place ecological and
evolutionary change on a shared theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\protect\hyperlink{ref-Turchin2001}{2001})
argues for a simple birth and death model (Box 1) as a fundamental
equation of population ecology, and for exponential growth to
consequently be as fundamental to ecology as Newton's first law of
motion is to physics. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental
(\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Frank2015a}{Frank 2017};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2020}{Lehtonen, Okasha, and Helanterä
2020}), and it serves as the starting point for all other fundamental
equations in evolution (\protect\hyperlink{ref-Queller2017}{Queller
2017}). We believe that this logical foundation can be made sufficiently
broad to encompass both population and evolutionary change.

@@

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\protect\hyperlink{ref-Price1970}{Price 1970},
\protect\hyperlink{ref-Price1972}{1972}), it postulates some basic
properties that all evolutionary systems must satisfy: change over time,
ancestor and descendant relations, and a character or phenotype
(\protect\hyperlink{ref-Rice2004}{Rice 2004}). Using simple algebraic
language, the Price equation represents evolutionary change with the
predominant notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Okasha2006}{Okasha 2006};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). A more specific version
of the covariance term was already known within the quantitative and
population genetics tradition
(\protect\hyperlink{ref-Robertson1966}{Robertson 1966}), usually
representing the action of natural selection. The Price equation adds an
expectation term and abstracts away from any specific mechanisms of
replication or reproduction, or mechanisms of inheritance. Its
definitional nature and lack of substantive biological assumptions has
been portrayed both as a strength (\protect\hyperlink{ref-Rice2004}{Rice
2004}; \protect\hyperlink{ref-Frank2012}{Frank 2012};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Baravalle2022}{Baravalle and Luque 2022}), and
its greatest weakness. The abstract nature of the Price equation places
it at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fischer's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\protect\hyperlink{ref-Queller2017}{Queller 2017};
\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
\protect\hyperlink{ref-Lehtonen2018}{2018}). This abstractness is also
key to developing a more general view of evolution
(\protect\hyperlink{ref-Rice2020}{Rice 2020};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
\protect\hyperlink{ref-Edelaar2023}{Edelaar, Otsuka, and Luque 2023}).
In contrast, some researchers consider the Price equation just a
triviality (even tautological), and useless without further modelling
assumptions (\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
\protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
debate remains open (\protect\hyperlink{ref-VanVeelen2020}{van Veelen
2020}; \protect\hyperlink{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations (e.g.,
\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\protect\hyperlink{ref-Page2002}{2002}) demonstrate a mathematical
equivalence between the Price equation and a general equation for
community dynamics. In both equations, the relative frequencies of
different ecological (species) or evolutionary (phenotype) variants map
to their respective ecological (population growth) or evolutionary
(fitness) change in a way that is mathematically equivalent. Collins and
Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use the Price
equation to build a general quantitative framework that partitions
phenotypic change into ecological, evolutionary, and physiological
components. The average change in a community-wide trait (e.g., carbon
uptake) is exactly and generally partitioned into change attributable to
changes in species composition, expected evolution of lineages within
species, and expected change in lineage physiology across all species.
Both Page and Nowak (\protect\hyperlink{ref-Page2002}{2002}) and Collins
and Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use relative
species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002}) or expected
change in a community-wide trait
(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but a
different approach is needed to recover absolute change in population
size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of
biological change (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Frank1997}{Frank 1997};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018};
\protect\hyperlink{ref-Queller2017}{Queller 2017}). It has also been
highly useful for integrating evolutionary theory across disciplines
(\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Brantingham2007}{Brantingham 2007};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Frank2015}{Frank 2015};
\protect\hyperlink{ref-Godsoe2021}{Godsoe et al. 2021};
\protect\hyperlink{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (e.g., \protect\hyperlink{ref-Page2002}{Page
and Nowak 2002}) or addition of terms (e.g.,
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
through the use of its recursive structure (e.g.,
\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). But despite its
flexibility, the Price equation still relies on relative frequencies,
which must by definition sum to one
(\protect\hyperlink{ref-Frank2015}{Frank 2015}). This is because the
Price equation describes the average change in a population; the
frequency of entities is scaled thereby conserving total probability
(\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}). But to recover the fundamental
principle of exponential population growth
(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}), this scaling must
be avoided in a fundamental equation of ecology and evolution.

@@

We therefore recover the general equation for population ecology (eqn 2)
and the fundamental property of exponential growth in populations (eqn
3, \protect\hyperlink{ref-Turchin2001}{Turchin 2001}).

@@

We have shown that we can derive the fundamental equations of population
ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining
ecosystem function and further conceptual unification between ecology,
evolution, and ecosystem function. The Price equation has previously
been used to investigate ecosystem function
(\protect\hyperlink{ref-Loreau2001}{Loreau and Hector 2001};
\protect\hyperlink{ref-Fox2006}{Fox 2006}), but not with any attempt
towards conceptual unification with evolutionary biology. For example,
Fox (\protect\hyperlink{ref-Fox2006}{2006}) applied the abstract
properties of the Price equation to partition total change in ecosystem
function into separate components attributable to species richness,
species composition, and context dependent effects. This approach
provides a framework for comparing the effects of biodiversity on
ecosystem function in empirical systems (e.g.,
\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Winfree2015}{Winfree et al. 2015};
\protect\hyperlink{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}).
Instead, our eqn 1 defines \(\Omega\) as total ecosystem function
contributed by a focal population (e.g., biomass production,
decomposition, photosynthetic rate). It is therefore possible to
investigate ecological, evolutionary, and ecosystem function change from
the same shared framework.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Biological evolution is realised through the same mechanisms of birth
and death that underlie change in population density. The deep
interdependence between ecology and evolution is well-established, but
much theory in each discipline has been developed in isolation. While
recent work has accomplished eco-evolutionary synthesis, a gap remains
between the logical foundations of ecology and evolution. We bridge this
gap with a new equation that defines a summed value for a characteristic
across individuals in a population, from which the fundamental equations
of population ecology and evolutionary biology (the Price equation) are
derived. We thereby unify the fundamental equations of population
ecology and biological evolution under a general framework. Our
unification further demonstrates the equivalence between mean population
growth rate and evolutionary fitness, shows how ecological and
evolutionary change are reflected in the first and second statistical
moments of fitness, respectively, and links this change to ecosystem
function. Finally, we outline how our proposed framework can be used to
unify social evolution and density-dependent population growth.

@@

@@

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\protect\hyperlink{ref-Kitcher1993}{Kitcher
1993}), and its value is perhaps most evident in reconciling unconnected
models and revealing new and unexpected empirical predictions. In
theoretical physics, Newton's unification of celestial and terrestrial
motion, and Maxwell's unification of electricity and magnetism, are
among the most important advances in scientific history. Likewise, in
the life sciences, the modern synthesis of natural selection and
Mendelian inheritance revolutionised how mechanisms underlying evolution
and genetics are understood and applied
(\protect\hyperlink{ref-Gayon2019}{Gayon and Huneman 2019}). The analogy
to physics is imperfect because evolution encompasses non-biological
systems with non-genetic inheritance
(\protect\hyperlink{ref-Price1995}{Price 1995};
\protect\hyperlink{ref-Shennan2011}{Shennan 2011};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Lewens2015}{Lewens 2015}), but to illustrate our
point, we restrict our focus to biological evolution. Following the
modern synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

@@

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (\protect\hyperlink{ref-Lion2018}{Lion 2018};
\protect\hyperlink{ref-Patel2018}{Patel et al. 2018};
\protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
foundational assumptions and equations of ecology and evolution in these
models remain mostly separated conceptually. But like the mechanisms
driving evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\protect\hyperlink{ref-Connor2004}{Connor and Hartl 2004}). Most
notably, change in population density and phenotype are both caused by
individual birth and death (\protect\hyperlink{ref-Turchin2001}{Turchin
2001}). This should allow biologists to place ecological and
evolutionary change on a shared theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\protect\hyperlink{ref-Turchin2001}{2001})
argues for a simple birth and death model (Box 1) as a fundamental
equation of population ecology, and for exponential growth to
consequently be as fundamental to ecology as Newton's first law of
motion is to physics. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental
(\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Frank2015a}{Frank 2017};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it
serves as the starting point for all other fundamental equations in
evolution (\protect\hyperlink{ref-Queller2017}{Queller 2017}). We
believe that this logical foundation can be made sufficiently broad to
encompass both population and evolutionary change.

@@

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\protect\hyperlink{ref-Price1970}{Price 1970},
\protect\hyperlink{ref-Price1972}{1972}), it postulates some basic
properties that all evolutionary systems must satisfy: change over time,
ancestor and descendant relations, and a character or phenotype
(\protect\hyperlink{ref-Rice2004}{Rice 2004}). Using simple algebraic
language, the Price equation represents evolutionary change with the
predominant notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Okasha2006}{Okasha 2006};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). A more specific version
of the covariance term was already known within the quantitative and
population genetics tradition
(\protect\hyperlink{ref-Robertson1966}{Robertson 1966}), usually
representing the action of natural selection. The Price equation adds an
expectation term and abstracts away from any specific mechanisms of
replication or reproduction, or mechanisms of inheritance. Its
definitional nature and lack of substantive biological assumptions has
been portrayed both as a strength (\protect\hyperlink{ref-Rice2004}{Rice
2004}; \protect\hyperlink{ref-Frank2012}{Frank 2012};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Baravalle2022}{Baravalle and Luque 2022}), and
its greatest weakness. The abstract nature of the Price equation places
it at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fischer's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
\protect\hyperlink{ref-Lehtonen2018}{2018};
\protect\hyperlink{ref-Queller2017}{Queller 2017}). This abstractness is
also key to developing a more general view of evolution
(\protect\hyperlink{ref-Rice2020}{Rice 2020};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
\protect\hyperlink{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast,
some researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
\protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
debate remains open (\protect\hyperlink{ref-VanVeelen2020}{van Veelen
2020}; \protect\hyperlink{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\protect\hyperlink{ref-Page2002}{2002}) demonstrate a mathematical
equivalence between the Price equation and a general equation for
community dynamics. In both equations, the relative frequencies of
different ecological (species) or evolutionary (phenotype) variants map
to their respective ecological (population growth) or evolutionary
(fitness) change in a way that is mathematically equivalent. Collins and
Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use the Price
equation to build a general quantitative framework that partitions
phenotypic change into ecological, evolutionary, and physiological
components. The average change in a community-wide trait (e.g., carbon
uptake) is exactly and generally partitioned into change attributable to
changes in species composition, expected evolution of lineages within
species, and expected change in lineage physiology across all species.
Both Page and Nowak (\protect\hyperlink{ref-Page2002}{2002}) and Collins
and Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use relative
species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002}) or expected
change in a community-wide trait
(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but a
different approach is needed to recover absolute change in population
size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of
biological change (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Frank1997}{Frank 1997};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Queller2017}{Queller 2017};
\protect\hyperlink{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
highly useful for integrating evolutionary theory across disciplines
(\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Brantingham2007}{Brantingham 2007};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Frank2015}{Frank 2015};
\protect\hyperlink{ref-Godsoe2021}{Godsoe et al. 2021};
\protect\hyperlink{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (\protect\hyperlink{ref-Page2002}{Page and
Nowak 2002}) or addition of terms
(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), or
through the use of its recursive structure
(\protect\hyperlink{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). But despite its
flexibility, the Price equation still relies on relative frequencies,
which must by definition sum to one
(\protect\hyperlink{ref-Frank2015}{Frank 2015}). This is because the
Price equation describes the average change in a population; the
frequency of entities is scaled thereby conserving total probability
(\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}). But to recover the fundamental
principle of exponential population growth
(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}), this scaling must
be avoided in a fundamental equation of ecology and evolution.

@@

We therefore recover the general equation for population ecology (eqn 2)
and the fundamental property of exponential growth in populations
(\protect\hyperlink{ref-Turchin2001}{Turchin 2001}) (eqn 3).

@@

We have shown that we can derive the fundamental equations of population
ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining
ecosystem function and further conceptual unification between ecology,
evolution, and ecosystem function. The Price equation has previously
been used to investigate ecosystem function
(\protect\hyperlink{ref-Loreau2001}{Loreau and Hector 2001};
\protect\hyperlink{ref-Fox2006}{Fox 2006}), but not with any attempt
towards conceptual unification with evolutionary biology. For example,
Fox (\protect\hyperlink{ref-Fox2006}{2006}) applied the abstract
properties of the Price equation to partition total change in ecosystem
function into separate components attributable to species richness,
species composition, and context dependent effects. This approach
provides a framework for comparing the effects of biodiversity on
ecosystem function in empirical systems
(\protect\hyperlink{ref-Fox2006}{Fox 2006};
\protect\hyperlink{ref-Winfree2015}{Winfree et al. 2015};
\protect\hyperlink{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}).
Instead, our eqn 1 defines \(\Omega\) as total ecosystem function
contributed by a focal population (e.g., biomass production,
decomposition, photosynthetic rate). It is therefore possible to
investigate ecological, evolutionary, and ecosystem function change from
the same shared framework.

@@

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

This work does not include any data.
